\vspace{1.0cm} \LARGE \bf Revealing the competitive dynamics of influenza viruses through evolutionary cartography \\ or \\ Charting antigenic drift and evolutionary success of seasonal influenza lineages

### 2013-02-01 flux.tex
Implement revisions suggested by Alan.

Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity to previous virus strains acquired by the host population.
Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
Here, we extend previous approaches to antigenic cartography, which seeks to place strains on an antigenic map, such that distances on this map best recapitulate titers observed across multiple HI assays.
In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
Using HI data for four lineages of influenza, encompassing influenza A subtypes H3N2 and H1N1, and influenza B lineages Victoria and Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
Through comparison with epidemiological data, we demonstrate a year-to-year correlation between drift and incidence and present evidence that antigenic drift mediates interference between influenza lineages.
We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences relative to other influenza lineages.
This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.

@@

Seasonal influenza infects between 10\% and 20\% of the human population every year, causing 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection. 
The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.

@@

We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza isolations attributable to each clade (see Methods).
We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of influenza isolations in the USA during this time period.
We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).

We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of virus isolations attributable to a clade for each season.
This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A). 
Here, within-clade drift explains 35\% of the variance in standardized incidence and presents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).

We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence in A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
We fit a linear model that predicts standardized incidence based upon drift within clades, drift in sister clades, and drift in other clades (Table~\ref{interferencetable}).
For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in sister clade A/H1N1 and by drift in other clades B/Vic and B/Yam.
We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of antigenic drift in a sister clade decreases incidence by 0.61 standard deviations and 1 unit of antigenic drift in other clades decreases incidence by 0.32 standard deviations.

@@

Furthermore, these findings suggest that the predominate clade of a particular influenza season could be estimated ahead of time by examining the degree of antigenic drift between influenza clades.
However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.

@@

Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving an advantage to influenza A in terms of overall rate of antigenic drift.
Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them an advantage over their sister lineages A/H1N1 and B/Yam also in terms of overall rate of antigenic drift.

@@

%%% incidence %%%
\begin{figure}[H]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/incidence}
	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of virus isolations of a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
	} 
	\label{incidence} 
\end{figure}

>>>>

Influenza viruses undergo continual antigenic evolution allowing mutant viruses to evade immunity acquired by the host population to previous virus strains.
Antigenic phenotype is often assessed through pairwise measurement of cross-reactivity between influenza strains using the hemagglutination inhibition (HI) assay.
Here, we extend previous approaches to antigenic cartography, which seeks to place strains on an antigenic map, such that distances on this map best recapitulate titers observed across multiple HI assays.
In our model, we simultaneously characterize antigenic and genetic evolution by including an evolutionary model in which antigenic location diffuses over a shared virus phylogeny.
Using HI data for four lineages of influenza, encompassing influenza A subtypes H3N2 and H1N1, and influenza B lineages Victoria and Yamagata, we determine average rates of antigenic drift for each lineage, as well as year-to-year variability in the rate of drift.
Through comparison with epidemiological data, we demonstrate a year-to-year correlation between drift and incidence and present evidence that antigenic drift mediates interference between influenza lineages.
We investigate the selective underpinnings for differing antigenic dynamics across lineages and show that A/H3N2 benefits from both a higher influx of new antigenic mutations and also from more efficient conversion of antigenic variation into fixed differences.
This work does much to elucidate the antigenic dynamics of influenza lineages, but also allows for substantial future advances in investigating the dynamics of influenza and other antigenically-variable pathogens by providing a model that intimately combines molecular and antigenic evolution.

@@

Seasonal influenza infects between 10\% and 20\% of the human population every year, causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}. 
While individuals develop long-lasting immunity to particular influenza strains after infection, antigenic mutations to the influenza virus genome result in proteins that are recognized to a lesser degree by the human immune system, leaving individuals susceptible to future infection. 
The influenza virus population continually evolves in antigenic phenotype in a process known as antigenic drift. 
A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective. 
A thorough understanding of the process of antigenic drift is essential to public health efforts to control mortality and morbidity through the use of a seasonal influenza vaccine.

@@

We investigate the relationship between rates of antigenic drift and USA incidence in A/H3N2, A/H1N1, B/Vic and B/Yam.
We take estimates of the rate of antigenic drift from the preceding section and compare these rates to measurements of relative incidence for each influenza clade calculated from the proportion of influenza viruses attributable to each clade (see Methods).
We analyze incidence from the 1998/1999 to the 2008/2009 seasons to avoid possible complications from the 2009 pandemic, finding that A/H3N2 accounted for 56\%, A/H1N1 for 20\%, B/Vic for 10\% and B/Yam for 14\% of viruses detected in the USA during this time period.
We find a significant correlation between rate of antigenic drift and relative incidence across the four clades  (Pearson correlation, $r=0.96$, $p = 0.021$).

We follow-up this analysis with a more detailed analysis of year-to-year variation in antigenic drift and clade-specific incidence in the USA.
We measure year-to-year antigenic drift within each clade as described in the previous section and shown in Figure \ref{jumps}.
We measure seasonal incidence of each clade by taking average influenza-like illness (ILI) percentage and multiplying this by the proportion of viruses attributable to a clade for each season.
This measure of incidence has previously been shown to have have predictive power in the analysis of seasonal influenza trends \cite{Goldstein11}.
In correlating antigenic drift to seasonal incidence we standardize this measure of incidence across seasons and across clades to measure effects in terms of standard deviations of incidence.
We find that years with high levels of antigenic drift coming into an influenza season, e.g.\ drift in A/H3N2 from 2000 to 2001, show high levels of incidence in that season, e.g.\ incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}A). 
Here, within-clade drift explains 35\% of the variance in standardized incidence and presents a significant correlation (Pearson correlation, $r=0.60$, $p=1.6 \times 10^{-5}$).

We further show that incidence within a clade decreases with antigenic drift in other clades, e.g.\ drift in A/H1N1, B/Vic and B/Yam from 2000 to 2001 decrease incidence of A/H3N2 in the 2001/2002 season (Figure~\ref{incidence}B).
This also represents a significant relationship (Pearson correlation, $r=-0.20$, $p=0.024$).
We fit a linear model that predicts standardized incidence based upon drift within clades, drift in sister clades, and drift in other clades (Table~\ref{interferencetable}).
For example, standardized incidence in A/H3N2 would be explained by drift within A/H3N2, drift in sister clade A/H1N1 and by drift in other clades B/Vic and B/Yam.
We compared adjusted $R^2$ across models with different regression coefficients included, and arrived at a best-fitting model that explains 46\% of the variance in standardized incidence and includes the terms $\bwithin$, $\bsister$ and $\bother$, but does not include an intercept term.
We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of antigenic drift in a sister clade decreases incidence by 0.61 standard deviations and 1 unit of antigenic drift in other clades decreases incidence by 0.32 standard deviations.

@@

Furthermore, these findings suggest that the predominant clade of a particular influenza season could be estimated ahead of time by examining the degree of variation in antigenic drift between influenza clades.
However, this prediction would not be perfect, and 56\% of the variance in incidence remains unexplained by these estimates of antigenic drift.
Some of this variance may be explained by incorporating geographic variation in year-to-year antigenic drift; although there exists significant geographic variation in clade-specific incidence \cite{Finkelman07}, we calculated antigenic drift from global samples, rather than USA-specific samples.

@@

Mutations must proceed through the sieve of natural selection, first appearing as transient differences restricted to side branches, while advantageous mutations will preferentially fix in the virus population and thus accrue on the trunk of the phylogeny.
We expect the diffusion coefficient on side branches to more closely resemble the underlying rate of antigenic mutation, before selection can strongly influence observed rates.
Here, we find that A/H3N2 and B/Vic have similar diffusion coefficients on side branches, which are larger than the diffusion coefficients of A/H1N1 and B/Yam (Table~\ref{diffusiontable}).
Thus, we expect that new antigenic mutations may occur at a faster rate in A/H3N2 and B/Vic than in A/H1N1 or B/Yam.
We find that trunk branches in both A/H3N2 and A/H1N1 evolve in antigenic phenotype faster than side branches (Table~\ref{diffusiontable}), suggesting that positive selection is promoting antigenic movement in these lineages.
Interestingly, B/Vic and B/Yam show less of a signature of positive selection, with similar diffusion coefficients on trunk and side branches (Table~\ref{diffusiontable}).
Thus, although drift forward along dimension 1 is driven by selection in all lineages (Figure~\ref{drift}), raw antigenic diffusion appears to be selected for more strongly in influenza A than in influenza B, giving influenza A an advantage in overall rate of antigenic drift.
Additionally, it appears that A/H3N2 and B/Vic benefit from a greater influx of new antigenic mutations, giving them a similar advantage over their sister lineages A/H1N1 and B/Yam, respectively.

@@

%%% incidence %%%
\begin{figure}[H]
	\centering		
	\includegraphics[width=0.95\textwidth]{figures/incidence}
	\caption{\textbf{Relationship between antigenic drift within and between clades and seasonal USA incidence for years 1998 to 2009.} 
	(A) Standardized incidence within a clade compared to antigenic drift within the same clade.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H3N2 from 1997 to 1998.
	(B) Standardized incidence within a clade compared to antigenic drift in other clades.
	For example, incidence of A/H3N2 in season 1998/1999 is measured against year-to-year antigenic drift of A/H1N1 from 1997 to 1998, as well as antigenic drift of B/Vic from 1997 to 1998.
	Standardized incidence is measured as yearly ILI percentage multiplied by proportion of viruses identified in a particular clade, and then standardized so that mean and variance across seasons and clades is equal to 0 and 1 respectively.
	Colors represent the clade in which incidence is measured; light blue for A/H3N2, purple for A/H1N1, orange for B/Vic and red for B/Yam.
	} 
	\label{incidence} 
\end{figure}
